Kindly Bent to Free Us

### 2019-07-03 kindly/motivation.tex
more motivation

\subsection{A session on linearity}

@@

To run client and server, we can create a channel and apply
\lstinline{add_service} to one end and \lstinline{op_client} to the other.
Failure to consume either channel endpoints (\lstinline/a/ or \lstinline/b/)
would result in a type error.
\begin{lstlisting}
let main1 () =
  let (a, b) = create () in
  fork add_service a;
  op_client b 1 2
# main : unit -> int
\end{lstlisting}

@@

\subsection{Imperative programming}

The programming style used for channels so far is
appropriate for purely functional code.
This style is however fairly bothersome in mutable contexts.
%
To make linearity easier to use for imperative programming, we rely on
{\em borrows}. We present a typical API for mutable affine arrays
in \cref{sig:array} and an example of use in \cref{ex:array}.
We can distinguish three type of operations. Firstly,
function that manipulate the whole
array such as \lstinline/create/ and \lstinline/free/, and behaves exactly
like the functions on channel presented in the previous section.
Second, read-only functions such as \lstinline/get/
manipulates \emph{immutable} borrows. An immutable borrow over array \lstinline/a/
is introduced with \lstinline/&a/ and represent a read-only view of the array.
Finally, mutable functions, such as \lstinline/set/ manipulate
a \emph{mutable borrow}. Mutable borrows, introduced with \lstinline/&!a/
represent a read-write view of the array.

Multiple immutable borrows can be used concurrently and are considered
unrestricted (\lstinline/un/).
A mutable borrow is exclusive, and considered affine (\lstinline/aff/).
Furthermore, any borrow prevents access to the ``raw'' access
as long as they are in use. In \cref{ex:array}, on \cref{ex:array:get},
two immutable borrows are used in the same expression. Once this operation
is done, the borrows are not used anymore, which allows us to take
an exclusive immutable borrow to mutate the array on \cref{ex:array:set}.
Finally we can return the whole array on \cref{ex:array:return} since
none of the borrows are used anymore.

@@

Naturally, \lang also checks that borrows do not escape their point
of use through {\em syntactic regions}. The example
above is reproduce in \cref{ex:array:region} with
regions explicitly delimited by \lstinline/{| ... |}/.
\lang ensures that borrows never escape the smallest enclosing
region.
For instance, On \cref{ex:array:region:get}, \lstinline/&a/ does
not escape the region since the borrow is not captured, and the
region only return integers.
Regions annotations are automatically inserted by \lang using
syntactic information such as borrows and binders.
Users can also annotate regions manually to further restrict the range
of a borrow. In practice, we have found that regions
annotations are very rarely needed in common imperative usage.

@@

\begin{figure}
  \centering
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
let for f a  = 
  let n = Array.length &a in
  let rec loop f a i = 
    if i < n then (
      f &!a i;
      loop f &!a (i+1)
    )
  in
  loop f &!a 0
# Borrowed m ('a Array.t) tt =>
  (tt -> int -> unit) -> tt -> unit
\end{lstlisting}
    \caption{For loop function for affine arrays}
    \label{ex:array}
  \end{subfigure}
\end{figure}

>>>>

\subsection{A session on linearity}
\label{sec:session-linearity}

@@

To run client and server, we can create a channel and apply
\lstinline{add_service} to one end and \lstinline{op_client} to the other.
Failure to consume either channel endpoints (\lstinline/a/ or \lstinline/b/)
would result in a type error.
\begin{lstlisting}
let main () =
  let (a, b) = create () in
  fork add_service a;
  op_client b 1 2
# main : unit -> int
\end{lstlisting}

@@

\subsection{Imperative programming}
\label{sec:imper-progr}

The programming style demonstrated for channels is
appropriate for purely functional code, but becomes tedious when
manipulating mutable resources.
%
To make linearity easier to use for imperative programming, we rely on
the notion of borrowing \cite{DBLP:conf/popl/BoylandR05}. As an example,
\cref{sig:array} contains an API for mutable affine arrays with a use case in \cref{ex:array}.

We distinguish several kinds of operation: constructors, destructore,
observers, and mutators. 
Constructors and destructors like \lstinline/create/ and \lstinline/free/ manipulate the whole
array. As with channels in \cref{sec:session-linearity},
\lstinline/create/ yields a linear resource which is consumed by \lstinline/free/.
During the lifetime of the array resource \lstinline/a/, we can split
off \emph{immutable borrows} \lstinline/&a/ that provide a read-only
view or \emph{mutable borrows} \lstinline/&!a/ for read-write views.
Observer functions such as \lstinline/get/ expect an immutable borrow
argument, mutator functions such a \lstinline/set/ expect
a mutable borrow.

Generally, each borrow is tied to a region whose lifetime is properly
contained in the lifetime of the resource. 
In a region, we can split off as many immutable borrows as we like
from a resource, but we can have only one mutable borrow. In a
subsidiary region, we can take further immutable borrows from any
borrow or we can take a mutable borrow from a mutable borrow from a
enclosing region. Borrows are confined to their regions. Inside,
immutable borrows are unrestricted (\lstinline/un/) whereas mutable
borrows are affine (\lstinline/aff/).

Furthermore, any borrow prevents access to the ``raw''  resource
as long as it is in use. In \cref{ex:array}, \cref{ex:array:get},
two immutable borrows are used in the same expression. Once this operation
is done, the borrows are not used anymore, which allows us to use
an exclusive mutable borrow in \cref{ex:array:set}.
Finally, we return the whole array in \cref{ex:array:return} because
none of the borrows are in scope anymore.

@@

%%% TBC
Naturally, \lang also checks that borrows do not escape their point
of use through {\em syntactic regions}. The example
above is reproduce in \cref{ex:array:region} with
regions explicitly delimited by \lstinline/{| ... |}/.
\lang ensures that borrows never escape the smallest enclosing
region.
For instance, On \cref{ex:array:region:get}, \lstinline/&a/ does
not escape the region since the borrow is not captured, and the
region only return integers.
Regions annotations are automatically inserted by \lang using
syntactic information such as borrows and binders.
Users can also annotate regions manually to further restrict the range
of a borrow. In practice, we have found that regions
annotations are very rarely needed in common imperative usage.

@@

\begin{figure}
  \centering
  \begin{subfigure}{0.5\linewidth}
\begin{lstlisting}
let for f a  = 
  let n = Array.length &a in
  let rec loop f a i = 
    if i < n then (
      f &!a i;
      loop f &!a (i+1)
    )
  in
  loop f &!a 0
# Borrowed m ('a Array.t) tt =>
  (tt -> int -> unit) -> tt -> unit
\end{lstlisting}
    \caption{For loop function for affine arrays}
    \label{ex:array:for}
  \end{subfigure}
\end{figure}
